\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage{natbib}
\usepackage{multirow}

\title{NepTTS-Bench: A Comprehensive Benchmark for\\Nepali Text-to-Speech Evaluation}
\author{Draft passing to collaborators}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Evaluating text-to-speech (TTS) for Nepali requires probing phonological contrasts that English-trained quality metrics cannot detect: a four-way laryngeal distinction (voiceless/voiced $\times$ unaspirated/aspirated) at five places of articulation, the retroflex--dental opposition, oral--nasal vowel contrasts, gemination, and schwa deletion. We present NepTTS-Bench, the first publicly available comprehensive benchmark for Nepali TTS. The benchmark is built around a phonologically-informed stimulus set of 77 minimal pairs targeting these contrasts within 205 sentences total. We evaluate 12 systems spanning commercial cloud services, Nepali-specific providers, open-source models, and human speech baselines. We collect 6{,}962 human MOS ratings from 207 native Nepali speakers (Krippendorff's $\alpha = 0.17$) and compare against 6 automated metrics, including SCOREQ auto-MOS and ASR round-trip CER across 5 ASR systems. Our key findings are: (1) the English-trained automated MOS predictor we test (SCOREQ) misranks systems relative to humans, putting ElevenLabs v3 first ($4.45$) while human raters rank it 6th of 12 ($3.47$); (2) Nepali-specific TTS systems are competitive with global cloud services, with TingTing Asmita the highest-rated TTS ($3.55$); (3) the choice of ASR drives Spearman $\rho$ between automated rankings and human MOS from $0.17$ (Whisper-small) to $0.60$ (Chirp~2), although none of these correlations reach significance at $\alpha = 0.05$ with $n = 9$ TTS systems; (4) our fine-tuned Nepali MOS predictor (NepaliMOS) reaches utterance-level Spearman $\rho = 0.59$ on held-out validation and system-level $\rho = 0.90$ ($p = 0.001$) against human MOS, significantly outperforming SCOREQ's $\rho = 0.40$ by Steiger's Z ($p = 0.028$). We release the complete benchmark as open-source resources, including the sentence sets, evaluation toolkit (\texttt{neptts-eval}), rating application, model checkpoint, and all data.
\end{abstract}

\section{Introduction}

Text-to-speech (TTS) technology has advanced dramatically, with neural models achieving near-human naturalness for high-resource languages. However, for low-resource languages such as Nepali, with an L1 plus L2 speaker population on the order of tens of millions across Nepal, India, Bhutan, and the global diaspora, systematic TTS evaluation remains absent. No standardized benchmark exists for comparing Nepali TTS systems, making it impossible for researchers and practitioners to assess progress or select appropriate systems.

Several systems now claim Nepali support: Microsoft Edge TTS, Google Gemini Flash, ElevenLabs v3, the Nepali-specific TingTing.io, and open-source Piper TTS. However, their relative quality has not been systematically compared by native speakers.

We present NepTTS-Bench with the following contributions:

\begin{enumerate}
\item The first Nepali MOS predictor (NepaliMOS), fine-tuned from IndicWav2Vec on 6{,}962 native-rater ratings. NepaliMOS reaches system-level Spearman $\rho = 0.90$ against human MOS (vs SCOREQ's $0.40$ on the same nine systems), significantly outperforming SCOREQ by Steiger's Z ($p = 0.028$).
\item A curated set of 205 sentences with 77 minimal pairs probing the major Nepali phonological contrasts (aspiration, retroflexion, nasalization, gemination, schwa deletion), supplemented with prosodic, lexical, and robustness probes (see Table~\ref{tab:categories}).
\item Evaluation of 12 systems (10 TTS + 2 human baselines) using 6 automated metrics and 6{,}962 human MOS ratings from 207 native speakers, with cluster-bootstrap confidence intervals and inter-rater reliability (Krippendorff's $\alpha = 0.17$) reported.
\item Evidence that two widely-used English-trained speech evaluation models fail on Nepali in distinct ways: SCOREQ for MOS prediction misranks the top systems relative to human raters, and Whisper large-v3-turbo for ASR hallucinates on $2\%$ to $10\%$ of files per system, with average CER reaching $7.92$ on gTTS. This motivates language-specific evaluation.
\item A complete open-source toolkit: benchmark sentences, evaluation CLI (\texttt{neptts-eval}), web-based rating application, and all evaluation data.\footnote{GitHub: \url{https://github.com/Ampixa/neptts-bench}, HuggingFace: \url{https://huggingface.co/datasets/ampixa/neptts-bench}}
\end{enumerate}


\section{Related Work}

\subsection{TTS Evaluation}
Mean Opinion Score (MOS) remains the gold standard for TTS evaluation \citep{streijl2016mos}. Automated MOS predictors include UTMOS \citep{saeki2022utmos}, which won the VoiceMOS Challenge 2022 \citep{huang2022voicemos}, SCOREQ \citep{ragano2024scoreq} from NeurIPS 2024, and DNSMOS \citep{reddy2022dnsmos}. However, these models are trained on English data and their cross-lingual reliability is not established for low-resource Indic languages.


\subsection{ASR Round-Trip Evaluation}
ASR round-trip evaluation measures intelligibility by transcribing synthesized speech and computing error rates against the original text. The TTSDS2 benchmark \citep{minixhofer2025ttsds} demonstrated that metric selection significantly affects system rankings. Our work extends this finding to a low-resource Indic setting, where the choice of ASR drives Spearman $\rho$ between automated rankings and human MOS from $0.17$ (Whisper-small) to $0.60$ (Chirp~2).

\subsection{Low-Resource Language TTS}
Meta's MMS-TTS \citep{pratap2023mms} covers 1,100+ languages including Nepali. AI4Bharat's Indic Parler-TTS supports Nepali with emotion-specific prompts. The FLEURS multilingual speech benchmark \citep{conneau2022fleurs} provides parallel utterances across 102 languages including Nepali but is oriented toward ASR and speech translation rather than TTS evaluation. No systematic TTS benchmark exists for evaluating Nepali TTS systems specifically.

\subsection{Nepali Speech Resources}
Nepali speech resources include the OpenSLR-54 corpus and fine-tuned Wav2Vec2/XLS-R ASR models \citep{gagan2023xlsr}. The strongest published Nepali ASR system to date is NepConformer \citep{poudel2026nepconformer}, a Conformer-based model trained on SLR54 that reports a Character Error Rate of $6.01\%$ and Word Error Rate of $23.96\%$ on the SLR54 dev set. The Chirp2 STT model from Google provides strong Nepali transcription. For TTS, no prior work has systematically evaluated available systems. To our knowledge, no publicly released benchmark combines (i) phonologically-informed Nepali stimuli, (ii) native-speaker MOS ratings at scale, (iii) multi-ASR round-trip evaluation, and (iv) an open evaluation toolkit. NepTTS-Bench is the first to do so.

\section{NepTTS-Bench Design}

\subsection{Sentence Set}

Our benchmark comprises 205 sentences designed to test specific aspects of Nepali TTS quality, informed by Nepali phonology \citep{khatiwada2009, pokharel1989, bandhu1971}. Table~\ref{tab:categories} summarizes the breakdown:

\begin{table}[h]
\centering
\caption{Sentence categories in NepTTS-Bench.}
\label{tab:categories}
\begin{tabular}{lcc}
\toprule
Category & Sentences & Pairs \\
\midrule
Aspiration (5 places) & 63 & 34 \\
Retroflex vs.\ dental & 20 & 10 \\
Oral vs.\ nasal vowel & 12 & 6 \\
Gemination & 16 & 8 \\
Schwa deletion & 18 & 9 \\
Nasal consonants & 20 & 10 \\
Question intonation & 10 & --- \\
Emotion & 8 & --- \\
Contrastive stress & 8 & --- \\
Homographs & 6 & --- \\
Newspaper text & 4 & --- \\
Robustness & 10 & --- \\
Phrases & 10 & --- \\
\midrule
\textbf{Total} & \textbf{205} & \textbf{77} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Minimal Pairs.} Nepali has a four-way laryngeal contrast (voiceless unaspirated, voiceless aspirated, voiced unaspirated, voiced aspirated) across five stop and affricate series at the bilabial, dental, retroflex, palatal, and velar places of articulation, a system absent in English \citep{khatiwada2009, pokharel1989}. For example, \textit{k\=am} (work) vs.\ \textit{kh\=am} (envelope) tests the velar aspiration contrast. Several caveats apply. The voiced aspirated (breathy) series is partially merged in casual speech, the retroflex--dental opposition tends to neutralize in coda position, and some contrasts (oral vs.\ nasal vowels, tatsama-driven gemination) carry lower functional load in Standard Nepali than they do in Hindi. Our stimulus set therefore probes the full phonological grid as it is written, and native raters may legitimately treat near-merged contrasts more leniently. We did not include breathy voice on sonorants (aspirated /m/, /n/, /l/), which a more comprehensive phonological audit would add. We designed 77 minimal pairs in which the two sentences differ by exactly one phonological feature. Each pair was audited for grammatical correctness and minimal pair validity. 12 pairs were replaced and 6 corrected during the audit process.

NepTTS-Bench is best understood as an engineering benchmark with phonologically-motivated stimuli rather than as a phonological study of Nepali in its own right. The minimal-pair design lets a TTS evaluator probe specific contrasts that English-trained quality predictors are unlikely to model, but a fully linguistic validation would require a trained phonetician's review of every pair and IPA-level transcription of all stimuli, which we leave to future work.

\paragraph{Additional Categories.} We include sentences testing question intonation (rising pitch patterns), emotional expression, contrastive stress, homograph disambiguation, robustness (tongue twisters, code-mixed text), and common phrases.

Additionally, 160 natural speech utterances from a Chirp2-transcribed Nepali YouTube corpus serve as human ceiling references.

\subsection{TTS Systems Evaluated}

We evaluate 10 TTS systems plus 2 human baselines (Table~\ref{tab:systems}), spanning commercial cloud services, Nepali-specific providers, and open-source models.

\begin{table}[h]
\centering
\caption{TTS systems evaluated in NepTTS-Bench.}
\label{tab:systems}
\begin{tabular}{llll}
\toprule
System & Provider & Type & Access \\
\midrule
ElevenLabs v3 & ElevenLabs & Cloud neural & API \\
Gemini Flash & Google & Cloud neural & API \\
Gemini (Hindi) & Google & Cloud neural & API \\
Edge TTS Hemkala & Microsoft & Cloud neural & Free \\
Edge TTS Sagar & Microsoft & Cloud neural & Free \\
TingTing Asmita & TingTing.io & Nepali-specific & API \\
TingTing Sambriddhi & TingTing.io & Nepali-specific & API \\
TingTing Subina & TingTing.io & Nepali-specific & API \\
Piper & Open source & Local VITS & Free \\
gTTS & Google Translate & Cloud & Free \\
\midrule
Human (read) & --- & Recorded & --- \\
Natural speech & YouTube & Spontaneous & --- \\
\bottomrule
\end{tabular}
\end{table}

Gemini Flash TTS auto-detects language from Devanagari text and tends to produce Hindi-accented pronunciation by default, since Devanagari is shared between Hindi and Nepali but Hindi dominates the model's training data. We therefore evaluate two prompting modes as separate ``systems''. \emph{Gemini (Hindi)} passes the raw Devanagari text with no language hint, eliciting the model's default Hindi-accented rendering. \emph{Gemini Flash} adds a prompt prefix that explicitly names Nepali as the target language, which steers the model toward Nepali pronunciation. Both use the \texttt{gemini-2.5-flash-preview-tts} model with voice ``Kore''.

\paragraph{Configuration disclosure.} The audio outputs we evaluate reflect the specific configurations listed in Table~\ref{tab:tts_configs}. Different voice IDs, model versions, prompt formulations, or generation parameters may yield meaningfully different results for the same provider; our findings should be read as a comparison of these specific configurations rather than as a verdict on each provider in general. We list every parameter we set so that a future evaluator can reproduce the audio set or quantify the effect of alternative settings.

\begin{table}[h]
\centering
\caption{TTS configurations evaluated. Output format and any deviation from API defaults are listed; unspecified parameters use the provider's defaults at the time of generation (April 2026).}
\label{tab:tts_configs}
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{p{0.18\linewidth}p{0.38\linewidth}p{0.36\linewidth}}
\toprule
System & Model / Voice & Configuration notes \\
\midrule
ElevenLabs v3 & \texttt{eleven\_multilingual\_v2}, voice ``Sarah'' & API defaults; ``Sarah'' is trained primarily on English. Voice ID in \texttt{scripts/generate\_elevenlabs\_tts.py}. MP3 output. \\
Gemini Flash & \texttt{gemini-2.5-flash-preview-tts}, voice ``Kore'' & Prompt prefix names Nepali explicitly. 24~kHz mono PCM. \\
Gemini (Hindi) & \texttt{gemini-2.5-flash-preview-tts}, voice ``Kore'' & Raw Devanagari, no language hint. 24~kHz mono PCM. \\
Edge TTS Hemkala & \texttt{ne-NP-HemkalaNeural} & Microsoft Edge TTS defaults. MP3 output. \\
Edge TTS Sagar & \texttt{ne-NP-SagarNeural} & Microsoft Edge TTS defaults. MP3 output. \\
TingTing Asmita & \texttt{np\_ashmita} & TingTing.io \texttt{test-voice} API. MP3 output. \\
TingTing Sambriddhi & \texttt{np\_sambriddhi} & TingTing.io \texttt{test-voice} API. MP3 output. \\
TingTing Subina & \texttt{np\_subina} & TingTing.io \texttt{test-voice} API. MP3 output. \\
Piper & Community Nepali Piper voice (VITS) & Local CPU inference, default sampling rate. \\
gTTS & \texttt{gTTS}, \texttt{lang="ne"} & Google Translate TTS endpoint. MP3 output. \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Audio characteristics.} The TTS-9 set comprises 1{,}729 utterances totaling $53.4$ minutes of audio (Table~\ref{tab:audio_stats}). File counts vary by system because some providers skip or fail on subsets of the 205 input sentences (notably Edge TTS and gTTS, which decline a small set of stimuli). Median utterance length differs substantially across systems, from $1.18$~s (TingTing Sambriddhi) to $2.23$~s (Edge TTS Hemkala), reflecting differences in speaking rate rather than content; all systems received the same 205 input sentences.

\begin{table}[h]
\centering
\caption{Audio duration statistics per TTS system over the 205 stimuli. All durations in seconds; \textit{Total} in minutes.}
\label{tab:audio_stats}
\footnotesize
\begin{tabular}{lrrrrrr}
\toprule
System & $n$ & Total & Median & Mean & Min & Max \\
\midrule
Edge TTS Hemkala       & 176 &  6.9 & 2.23 & 2.35 & 1.32 & 7.46 \\
Edge TTS Sagar         & 176 &  6.8 & 2.21 & 2.31 & 1.39 & 6.98 \\
gTTS                   & 176 &  6.7 & 2.11 & 2.28 & 0.74 & 8.42 \\
Gemini Flash           & 201 &  7.0 & 2.01 & 2.10 & 0.69 & 7.41 \\
Piper                  & 191 &  5.3 & 1.51 & 1.67 & 0.44 & 6.93 \\
TingTing Asmita        & 202 &  4.9 & 1.33 & 1.46 & 0.26 & 5.59 \\
TingTing Sambriddhi    & 202 &  4.4 & 1.18 & 1.31 & 0.21 & 5.96 \\
TingTing Subina        & 202 &  5.4 & 1.44 & 1.60 & 0.34 & 5.62 \\
ElevenLabs v3          & 203 &  6.0 & 1.72 & 1.78 & 0.52 & 6.27 \\
\midrule
\textbf{TTS-9 combined} & \textbf{1{,}729} & \textbf{53.4} & \textbf{1.72} & \textbf{1.85} & \textbf{0.21} & \textbf{8.42} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Evaluation Methodology}

\paragraph{Human MOS.} Native Nepali speakers rate blinded audio samples on a 1--5 scale via a web application. Raters choose their available time (5--20 minutes) and receive a balanced, randomized set of stimuli across all systems. Same-sentence stimuli are spaced at least 3 positions apart to reduce fatigue. The system identity is hidden.

\paragraph{ASR Round-Trip.} We transcribe all TTS outputs with 5 ASR systems and compute Character Error Rate (CER) against normalized Devanagari reference text:
\begin{itemize}
\item \textbf{Google Chirp~2}: Speech-to-Text v2 API, model \texttt{chirp\_2}, language code \texttt{ne-NP}.
\item \textbf{Meta MMS-1b-all}: \texttt{facebook/mms-1b-all} via Hugging Face transformers with the Nepali language adapter.
\item \textbf{Wav2Vec2 XLS-R Nepali}: \texttt{gagan3012/wav2vec2-xlsr-nepali} with greedy CTC decoding.
\item \textbf{Whisper-small}: OpenAI \texttt{whisper} model \texttt{small}, \texttt{language="ne"}, \texttt{task="transcribe"}, fp32 inference on GPU.
\item \textbf{Whisper large-v3-turbo}: via MLX (Apple Silicon port) with the same Nepali language hint.
\end{itemize}
Reference and hypothesis are NFC-normalized and stripped of common punctuation (Devanagari danda U+0964 and Latin punctuation) before edit-distance scoring. The Whisper-small evaluation script is at \texttt{scripts/eval\_asr\_roundtrip.py}; per-system raw transcriptions and CER values for all five ASRs are released as JSON in \texttt{benchmark/results/}.

\paragraph{Automated MOS.} We use SCOREQ \citep{ragano2024scoreq}, a NeurIPS 2024 model for no-reference speech quality assessment in the synthetic speech domain, using ONNX inference on CPU.

\paragraph{Pair Discrimination.} For minimal pairs, raters hear both sentences from a pair for a given TTS system and judge whether the contrast words sound different (Yes/No).

\paragraph{Spearman Rank Correlation.} We report Spearman's rank correlation coefficient ($\rho$) throughout to quantify agreement between rankings: $\rho = +1$ indicates perfect monotonic agreement, $0$ no relationship, and $-1$ perfect disagreement. Unlike Pearson correlation, $\rho$ operates on ranks rather than raw values. This makes it invariant to the absolute scale of the underlying scores, which is appropriate when comparing metrics with different ranges (e.g., MOS in $[1,5]$ vs.\ CER as a percentage). We follow the convention that $|\rho| < 0.3$ is weak, $0.3$--$0.6$ moderate, $0.6$--$0.8$ strong, and $>0.8$ very strong.

\paragraph{Statistical Inference.} Confidence intervals in Table~\ref{tab:human_mos} are 95\% cluster-bootstrap intervals computed by resampling raters with replacement (2{,}000 resamples), which accounts for within-rater dependence given the 34 ratings per rater on average. The resulting intervals are roughly twice as wide as naive Wald intervals would be, and they reveal that the top four TTS systems (TingTing Asmita, Piper, TingTing Subina, ElevenLabs v3) all have overlapping CIs. The qualitative ranking is robust at the gross level (gTTS clearly bottom, Nepali-specific and open-source systems near the top), but the ordering within the top four cannot be statistically distinguished and should not be over-interpreted.

\paragraph{Inter-Rater Reliability.} We compute Krippendorff's $\alpha$ on the rater-by-sample matrix using interval distance and obtain $\alpha = 0.17$ (ordinal $\alpha = 0.16$). This is well below the conventional $0.667$ threshold for tentative conclusions and below the $0.4$--$0.6$ range typical of well-controlled lab MOS studies. The rating task is therefore noisy: individual ratings disagree substantially across raters on the same audio, although per-system mean MOS still differentiates clearly among system tiers. We attribute the low $\alpha$ to the unscreened web-app rater pool, the absence of headphone or environment requirements, and the inherent difficulty of MOS as an absolute-scale judgment. Future revisions should report $\alpha$ alongside more controlled listening conditions or move toward forced-choice comparative ratings (CMOS).

\section{Results}

\subsection{Human MOS}

\begin{table}[t]
\centering
\caption{Human MOS ratings from 207 native Nepali speakers (6,962 total ratings). Means and 95\% confidence intervals are computed by cluster bootstrap with 2{,}000 resamples of raters with replacement. The intervals are wider than naive Wald intervals because each rater contributes an average of 34 ratings.}
\label{tab:human_mos}
\begin{tabular}{lccc}
\toprule
System & MOS & 95\% CI & Ratings \\
\midrule
Natural speech & \textbf{3.95} & $[3.79, 4.15]$ & 597 \\
Human (read) & 3.64 & $[3.44, 3.82]$ & 552 \\
\midrule
TingTing Asmita & 3.55 & $[3.27, 3.88]$ & 651 \\
Piper & 3.50 & $[3.31, 3.69]$ & 515 \\
TingTing Subina & 3.47 & $[3.24, 3.79]$ & 644 \\
ElevenLabs v3 & 3.47 & $[3.28, 3.66]$ & 516 \\
Edge TTS Sagar & 3.38 & $[3.02, 3.77]$ & 609 \\
Edge TTS Hemkala & 3.37 & $[3.10, 3.67]$ & 619 \\
Gemini (Hindi) & 3.34 & $[3.15, 3.53]$ & 514 \\
Gemini Flash & 3.25 & $[2.90, 3.68]$ & 618 \\
TingTing Sambriddhi & 3.20 & $[2.91, 3.55]$ & 617 \\
gTTS & 2.60 & $[2.38, 2.83]$ & 510 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:human_mos} presents the main human evaluation results. Key observations:

\textbf{Human ceiling.} Natural speech scores 3.95, not 5.0, reflecting that even human recordings have quality variation (background noise, recording equipment). Read speech (3.64) scores lower than spontaneous speech (3.95), possibly due to the unnatural cadence of prompted reading.

\textbf{Nepali-specific systems compete.} TingTing Asmita (3.55) is the highest-rated TTS system, ahead of global cloud services including ElevenLabs v3 (3.47) and both Edge TTS voices (3.37--3.38). The CIs of TingTing Asmita, Piper, TingTing Subina, and ElevenLabs all overlap, so the ordering among the top four cannot be statistically separated, but as a group they are clearly above the Edge TTS and Gemini variants.

\textbf{Open-source is competitive.} Piper (3.50), a fully open-source local model, places among the top-four TTS cluster.

\textbf{gTTS is unusable.} Google Translate TTS scores 2.60, far below all other systems and outside any other system's CI, confirming that concatenative TTS is inadequate for Nepali.

\subsection{Automated Metrics}

Table~\ref{tab:auto_metrics} reports SCOREQ auto-MOS and ASR round-trip CER for our 10 TTS systems across four of the five ASRs we evaluated. Whisper large-v3-turbo is excluded from the per-system table; we discuss its failure mode in Section~\ref{sec:whisper-large}.

\begin{table}[t]
\centering
\caption{Automated evaluation of 9 TTS systems. MOS metrics ($\uparrow$, higher is better) and CER metrics ($\downarrow$, lower is better). Whisper here is Whisper-small. Gemini (Hindi) is omitted because automated metrics were computed only on the Nepali-prompted Gemini Flash variant.}
\label{tab:auto_metrics}
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lcccccc}
\toprule
System & SCOREQ & NepaliMOS & Chirp2 & MMS & XLS-R & Whisper \\
 & $\uparrow$ & $\uparrow$ & CER $\downarrow$ & CER $\downarrow$ & CER $\downarrow$ & CER $\downarrow$ \\
\midrule
ElevenLabs v3 & \textbf{4.45} & 3.46 & \textbf{0.106} & 0.278 & 0.222 & 0.470 \\
TingTing Subina & 4.27 & 3.36 & 0.120 & \textbf{0.181} & \textbf{0.160} & 0.514 \\
TingTing Asmita & 4.17 & 3.40 & 0.141 & 0.234 & 0.198 & 0.484 \\
Gemini Flash & 4.10 & 3.20 & 0.144 & 0.214 & 0.199 & \textbf{0.372} \\
TingTing Sambriddhi & 4.05 & 3.18 & 0.216 & 0.313 & 0.234 & 0.554 \\
Edge TTS Hemkala & 3.74 & 3.22 & 0.170 & 0.301 & 0.161 & 0.457 \\
Edge TTS Sagar & 3.63 & 3.20 & 0.146 & 0.279 & 0.148 & 0.652 \\
Piper & 3.49 & \textbf{3.52} & 0.186 & 0.291 & 0.168 & 0.471 \\
gTTS & 3.02 & 2.51 & 0.473 & 0.531 & 0.279 & 0.586 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Human vs.\ Automated Divergence}

The most striking finding is the disagreement between automated and human metrics. Table~\ref{tab:divergence} shows the rank comparison.

\begin{table}[t]
\centering
\caption{Ranking divergence: automated SCOREQ vs.\ human MOS. Both ranks are computed over the nine TTS systems with SCOREQ scores (Gemini (Hindi) is excluded because no SCOREQ score was computed for it).}
\label{tab:divergence}
\begin{tabular}{lccc}
\toprule
System & SCOREQ & Human & $\Delta$Rank \\
 & Rank & Rank & \\
\midrule
ElevenLabs v3 & 1 & 4 & $-3$ \\
TingTing Subina & 2 & 3 & $-1$ \\
TingTing Asmita & 3 & \textbf{1} & $+2$ \\
Gemini Flash & 4 & 7 & $-3$ \\
TingTing Sambriddhi & 5 & 8 & $-3$ \\
Edge TTS Hemkala & 6 & 6 & $0$ \\
Edge TTS Sagar & 7 & 5 & $+2$ \\
Piper & 8 & \textbf{2} & $+6$ \\
gTTS & 9 & 9 & $0$ \\
\bottomrule
\end{tabular}
\end{table}

ElevenLabs v3 scores highest on SCOREQ (4.45) but humans rank it 4th among the nine TTS systems with SCOREQ scores ($\Delta\text{Rank} = -3$), or 6th of 12 overall when human baselines are included (3.47 MOS). The v3 model supports Nepali natively but uses an English voice (Sarah), producing technically clean audio with non-native pronunciation that automated metrics cannot detect. Conversely, Piper ranks 8th on SCOREQ but 2nd with humans ($\Delta\text{Rank} = +6$). Its pronunciation is more Nepali-native despite lower audio fidelity.

\subsection{Whisper Large-v3-Turbo Hallucinates on Nepali}
\label{sec:whisper-large}

We additionally evaluated Whisper large-v3-turbo (via MLX) as our fifth ASR. Its outputs are dominated by sporadic but extreme hallucinations: across the nine TTS systems, between $2\%$ and $10\%$ of files have $\text{CER} > 1.0$ (the hypothesis is longer than the reference), with maximum CER reaching $186$ on Edge TTS Hemkala and $1{,}115$ on gTTS. Average CER is therefore highly skewed: $7.92$ on gTTS and $2.29$ on Edge TTS Hemkala, against $0.22$ to $0.64$ on the remaining systems. Median CER, by contrast, falls in $0.15$ to $0.44$ on every system, comparable to or worse than Whisper-small. Manual inspection confirms repeated tokens and inserted phrases unrelated to the input audio in the high-CER files.

This is a meaningful negative finding. Whisper large-v3-turbo is widely used as a default ASR in research pipelines, and its hallucination behavior on Nepali extends beyond the simple unreliability we observe for Whisper-small ($\rho = 0.38$ in Table~\ref{tab:correlations}). Per-system averages computed naively over hallucinated runs are misleading. For low-resource Nepali TTS evaluation, Whisper large-v3-turbo should be used only with hallucination filtering. We exclude it from per-system reporting and recommend that future low-resource speech evaluations characterize ASR hallucination rate before relying on round-trip CER.

\subsection{Cross-Metric Correlations}

Table~\ref{tab:correlations} reports pairwise Spearman correlations among the five automated metrics computed across $n = 9$ TTS systems. Asterisks mark pairs that survive Benjamini--Hochberg correction at FDR $0.05$.

\begin{table}[t]
\centering
\caption{Spearman rank correlation ($\rho$) between automated metrics across $n = 9$ TTS systems. * indicates Benjamini--Hochberg-adjusted $p < 0.05$. CER metrics are negated so that higher values mean ``more agreement with the others''.}
\label{tab:correlations}
\begin{tabular}{lccccc}
\toprule
 & SCOREQ & Chirp2 & MMS & XLS-R & Whisper \\
\midrule
SCOREQ & 1.000 & 0.883* & 0.733 & 0.050 & 0.383 \\
Chirp2 & & 1.000 & 0.867* & 0.350 & 0.367 \\
MMS & & & 1.000 & 0.450 & 0.350 \\
XLS-R & & & & 1.000 & $-$0.033 \\
Whisper & & & & & 1.000 \\
\bottomrule
\end{tabular}
\end{table}

SCOREQ, Chirp~2, and MMS form a correlated cluster (all pairwise $\rho \geq 0.73$, with SCOREQ--Chirp~2 and Chirp~2--MMS surviving BH correction at FDR $0.05$). Whisper-small and XLS-R Nepali sit outside this cluster, with weak or no correlation to any other metric ($\rho \leq 0.45$).

\paragraph{Agreement with human MOS.} The cross-metric table above shows automated agreement, not agreement with human ratings. Table~\ref{tab:metric_vs_human} reports each metric's Spearman correlation against human MOS over the same nine TTS systems.

\begin{table}[h]
\centering
\caption{Spearman correlation between each automated metric and human MOS, computed across the nine TTS systems with SCOREQ scores. NepaliMOS, our Nepali-specific predictor, is the only metric whose correlation with human MOS is statistically significant at $\alpha = 0.05$.}
\label{tab:metric_vs_human}
\begin{tabular}{lcc}
\toprule
Metric & $\rho$ vs Human & Raw $p$ \\
\midrule
\textbf{NepaliMOS (ours)} & \textbf{0.90} & \textbf{0.001} \\
Chirp~2 & 0.60 & 0.09 \\
MMS & 0.58 & 0.10 \\
XLS-R Nepali & 0.53 & 0.14 \\
SCOREQ & 0.40 & 0.29 \\
Whisper-small & 0.17 & 0.67 \\
\bottomrule
\end{tabular}
\end{table}

The agreement-with-humans story has two parts. First, NepaliMOS, our Nepali-specific predictor (Section~\ref{sec:nepalimos}), reaches $\rho = 0.90$ with human ratings and is significantly better than SCOREQ at predicting human MOS by Steiger's Z ($Z = +2.19$, two-sided $p = 0.028$). This is the validation that the predictor is necessary: an English-trained predictor (SCOREQ) gives $\rho = 0.40$ on the same nine systems, while a Nepali-trained predictor on the same task gives $\rho = 0.90$. Second, among the English-trained metrics, Chirp~2 ranks systems most consistently with humans ($\rho = 0.60$) but the difference between Chirp~2 and Whisper-small is not statistically significant ($Z = 1.09$, $p = 0.28$). XLS-R Nepali, fine-tuned on Nepali ASR data, gives a different ranking from SCOREQ ($\rho = 0.05$ between them); we report this as observed but do not interpret it given the wide confidence interval at $n = 9$.

\subsection{NepaliMOS Predictor}
\label{sec:nepalimos}

\paragraph{Motivation.} Section~\ref{sec:scoreq-unreliable-anchor} shows that SCOREQ, the only English-trained automated MOS predictor we evaluate, misranks systems relative to native human raters and reaches only $\rho = 0.40$ with human MOS at the system level. A Nepali-specific predictor that learns from native-rater data should be able to do better. We train such a predictor on the 6{,}962 native-rater ratings collected via our rating application and evaluate it on both held-out utterances (the standard MOS-predictor evaluation) and held-out TTS systems (the evaluation that matters for ranking).

\paragraph{Architecture and training.} We fine-tune IndicWav2Vec base \citep{javed2022indicwav2vec} on the 6{,}962 ratings aggregated to 1{,}696 utterance-level mean-MOS targets. The top 4 transformer layers of the backbone are unfrozen and a 2-layer MLP regression head ($768 \rightarrow 256 \rightarrow 1$, with ReLU and dropout $0.1$) sits on top of mean-pooled hidden states. Training uses MSE loss with Adam at differential learning rates (backbone $10^{-5}$, head $10^{-4}$), batch size 2, and a stratified 90/10 utterance-level train/validation split by MOS bucket. We select the checkpoint by best validation loss, which lands at epoch 7 of 30. Both backbone and head weights are saved together so the artifact reproduces validation performance from the saved file alone.

\paragraph{Results.} On the held-out validation utterances ($n \approx 170$) we obtain utterance-level Spearman $\rho = 0.59$ and Pearson $r = 0.57$. At the system level, where we average NepaliMOS predictions over all of a system's audio files and correlate the resulting nine system-level means with human MOS, we obtain $\rho = 0.90$ ($p = 0.001$). The system-level number is what matters for ranking TTS systems and is the stronger validation: NepaliMOS more than doubles SCOREQ's $\rho = 0.40$ on the same set of nine systems and is the only automated metric whose system-ranking agreement with humans reaches statistical significance at $\alpha = 0.05$ (Table~\ref{tab:metric_vs_human}). Steiger's Z confirms NepaliMOS outperforms SCOREQ at predicting human MOS at $p = 0.028$. The split is utterance-disjoint but not system-disjoint, so leave-one-system-out cross-validation remains the appropriate generalization test and is left to future work.

Performance scales with data volume (Table~\ref{tab:scaling}):

\begin{table}[h]
\centering
\caption{NepaliMOS predictor scaling with training data. Numbers are from an earlier training snapshot; the trend is the relevant signal.}
\label{tab:scaling}
\begin{tabular}{ccc}
\toprule
Ratings & Samples & Spearman $\rho$ \\
\midrule
1,166 & 1,166 & 0.195 \\
1,430 & 1,430 & 0.305 \\
2,742 & 1,541 & 0.413 \\
5,760 & 1,678 & 0.587 \\
7,003 & 1,696 & \textbf{0.660} \\
\bottomrule
\end{tabular}
\end{table}

The trend in Table~\ref{tab:scaling} suggests that more ratings continue to improve $\rho$ at this data scale. We do not extrapolate a specific target since five points are insufficient to fit a calibrated curve, and we did not vary architecture in this experiment so we cannot conclude that data is the dominant bottleneck.

\section{Discussion}

\subsection{SCOREQ Is Unreliable for Nepali}
\label{sec:scoreq-unreliable-anchor}
SCOREQ, trained on English speech quality data, measures audio fidelity (signal clarity, noise levels) but not language-appropriate pronunciation. A system can produce technically perfect audio with an English accent on Nepali text and score high on automated metrics while sounding unnatural to native speakers. This gap motivates language-specific MOS predictors like our NepaliMOS.

\subsection{Nepali-Specific Systems Are Competitive}
TingTing.io, a Nepal-based TTS provider, achieves the highest human MOS among TTS systems (3.55). This demonstrates that language-specific training can compensate for smaller model scale, and has practical implications for applications targeting Nepali speakers.

\subsection{Open-Source Quality}
Piper TTS (3.50) is fully open-source and runs locally without cloud dependency. Its competitive performance, statistically indistinguishable from ElevenLabs and Edge TTS within our CIs, makes it viable for applications where privacy, cost, or connectivity constraints preclude cloud services.

\subsection{ASR Model Selection}
Across the five ASRs we tested, the Spearman correlation with human MOS ranges from $\rho = 0.17$ (Whisper-small) to $\rho = 0.60$ (Chirp~2), with MMS at $0.58$ and XLS-R Nepali at $0.53$ (Table~\ref{tab:metric_vs_human}). With $n = 9$ TTS systems, none of these correlations reach significance at $\alpha = 0.05$ and Steiger's Z test confirms that we cannot statistically distinguish Chirp~2's agreement with humans from Whisper-small's ($Z = 1.09$, $p = 0.28$). The differences are nonetheless practically meaningful: Chirp~2 places ElevenLabs and gTTS in roughly the right human-rank slots, while Whisper-small does not. Combined with Whisper large-v3-turbo's hallucination behavior (Section~\ref{sec:whisper-large}), our recommendation for Nepali TTS evaluation is: Chirp~2 or MMS as the primary ASR, with Whisper avoided unless paired with hallucination filtering.

\subsection{Limitations}

\paragraph{Rater pool and demographics.} Our rater pool of 207 speakers may not fully represent Nepal's linguistic diversity in geography, dialect, or L2 status. We treat ``native Nepali speaker'' as self-declared and do not collect demographic metadata in this version. The benchmark covers Standard Nepali stimuli only.

\paragraph{Listening environment.} Raters used the web rating application from their own devices and we do not enforce headphones, ambient noise control, or volume calibration. Standard ITU-T P.808 crowd-MOS protocols would screen for these; this is a meaningful threat to the absolute MOS values, although the relative ranking should be more robust.

\paragraph{Rater fatigue and order effects.} We space same-sentence stimuli at least three positions apart and randomize the within-rater stimulus order, but we do not analyze MOS drift across position-in-session. A non-trivial fraction of raters complete only the shorter session option (5 to 10 minutes) and longer sessions may exhibit rating compression toward the mean.

\paragraph{Same-system bias.} TTS systems are blinded to raters by stripping system identifiers, but voice characteristics (pitch, accent, prosody) are themselves identifying. Raters who learn to recognize a voice may anchor subsequent ratings of that voice. We do not currently audit this.

\paragraph{Statistical inference and inter-rater reliability.} Confidence intervals in Table~\ref{tab:human_mos} are 95\% cluster-bootstrap intervals over raters and the Krippendorff's $\alpha$ on the rater-by-sample matrix is $0.17$. The low $\alpha$ is itself a finding: web-app crowd-MOS without environment screening or rater training yields noisy individual ratings, even though per-system means still differentiate clearly among system tiers. Forced-choice comparative MOS (CMOS) and ITU-T P.808-style screening would likely improve $\alpha$ substantially.

\paragraph{NepaliMOS predictor.} The predictor reaches utterance-level $\rho = 0.66$ on a held-out validation split that is utterance-disjoint but not system-disjoint. Leave-one-system-out cross-validation is the appropriate generalization test and is left to future work.

\paragraph{Pair discrimination.} The pair-discrimination subtask has only 54 total ratings (under one rating per minimal pair on average), so we do not draw statistical conclusions from it. We retain it in the rating application for ongoing data collection.

\paragraph{Automated MOS scope.} We did not run UTMOS or DNSMOS, so the automated-MOS unreliability finding is established only for SCOREQ in our experiments. Whether the failure generalizes to other English-trained predictors is an open question.

\paragraph{Configuration sensitivity.} Several TTS configurations in Table~\ref{tab:tts_configs} reflect the specific settings we used. Alternative voice IDs, prompts, or generation parameters may yield meaningfully different results for the same provider; our ranking claims are therefore conditional on these configurations and not on the providers in general.

\section{Conclusion}

We presented NepTTS-Bench, the first publicly available comprehensive benchmark for Nepali TTS evaluation, with 205 sentences, 77 minimal pairs, 12 systems evaluated by 207 native speakers (6{,}962 ratings, Krippendorff's $\alpha = 0.17$), 6 automated metrics, and NepaliMOS, a Nepali-specific MOS predictor that reaches system-level Spearman $\rho = 0.90$ ($p = 0.001$) against human MOS, statistically significantly better than SCOREQ's $\rho = 0.40$ by Steiger's Z ($p = 0.028$). Our two key findings are that (i) SCOREQ, the only English-trained automated MOS predictor we evaluate, misranks systems relative to human raters; and (ii) a small Nepali-specific predictor, fine-tuned from IndicWav2Vec on the same human ratings, recovers the human ranking. Whether the SCOREQ result generalizes to UTMOS, DNSMOS, and other English-trained predictors is an open question. The Krippendorff's $\alpha$ of $0.17$ at the individual-rating level limits the precision with which the top four TTS systems can be ordered and motivates more controlled listening environments and forced-choice rating paradigms in low-resource TTS development.

All resources are publicly available: the benchmark sentence set, evaluation toolkit (installable via \texttt{pip install neptts-eval}), rating application, model checkpoint, and evaluation data.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
